---
title: A quick method for thousands of ANOVA for a common grouping variable in R
author: Junli Zhang
date: '2018-09-02'
slug: a-quick-method-for-thousands-of-anova-for-a-common-grouping-variable-in-r
categories:
  - R
tags:
  - Tips
---

I have a RNAseq data for two lines (R and W) at 3 collection point (day 3, 6, 9). I wanted to do ANOVAs to test whether the expression of each of the 100,000 genes is different between the two lines at each day. I did the ANOVAs for each day, now I wanted to the ANOVAs with contrasts using all the 6 treatment together (R3, R6, R9 and W3, W6, W9).

I found the code below is too slow, and it took about 3 hours to calculate the P values of all the genes.

```{r, eval=FALSE}
# dd7 is the organized data with one grouping factor "lineday" and the counts data of all the 100,000 genes.

# function to calculate pvalues of one gene
pvalue = function(marker){
  ff = as.formula(paste0(marker, "~lineday"))
  model = aov(ff, dd7)
  dum = summary.aov(model, split=list(lineday = list("day3"= 1, "day6" = 2, "day9"= 3)))
  return(dum[[1]][2:4,5]) # pvalues for day3, day6, day9
}
# contrasts used in the anova for lineday
day3  = c(1, 0,  0,  -1,  0,  0) # R3 vs. W3
day6  = c(0,  1, 0,  0,  -1,  0) # R6 vs. W6
day9  = c(0, 0, 1,  0,  0,  -1) # R9 vs. W9

mat = cbind(day16, day6, day9)
contrasts(dd7$lineday) <- mat # set the contrast before anova for lineday
nm = names(dd7) # first name is lineday
allp = sapply(nm[-1], pvalue) # pvalues for all markers
```
This is too slow, so I did a search and found [this post](http://r.789695.n4.nabble.com/Is-there-a-fast-way-to-do-several-hundred-thousand-ANOVA-tests-td900181.html). Hadley Wickham pointed to another article "Computing Thousands of Test Statistics Simultaneously in R" included in this R [newsletter]( http://stat-computing.org/newsletter/issues/scgn-18-1.pdf). And I found R builtin functions "lm" and "aov" can fit all the columns of y if y is a matrix, and I also found "summary" function can directly give you the F test results from the lm fits. Below is the code I updated and now it finishes in 1 minute.

```{r, eval=FALSE}
lineday = dd7[, 1] # get the grouping variable
contrasts(lineday) <- mat # set the contrast
system.time(mm <- aov(as.matrix(dd7[-1])~ll)) # only 1.3 sec

# below only takes 33 sec
system.time(ss  <-  summary.aov(mm, split=list(lineday = list("day3"= 1, "day6" = 2, "day9"= 3)))) # get all the summaries: a list of anovas for each gene
names(ss) = sub(" Response ", "", names(ss)) # change the names
allp = sapply(ss, function(x) x[2:4, 5]) # get all the pvalues
class(allp)
allp = t(allp) # transpose it
```

